\huge The human visual system preserves the hierarchy of 2-dimensional pattern regularity

### 2021-05-13 Writeup/kohler_clarke_2021.tex
addressed R2 comments, and two of the editor queries

Symmetries are abundant in natural and man-made environments, due to a complex interplay of physical forces that govern pattern formation in nature. Symmetrical patterns have been created and appreciated by human cultures throughout history and since the gestalt movement of the early 20th century, symmetry has been recognized as important for visual perception. Symmetry contributes to the perception of shapes \citep{RN1311,RN1682}, scenes \citep{RN1824} and surface properties \citep{RN1166}, as well as the social process of mate selection \citep{RN1337}. Most of this work has focused on mirror symmetry or \textit{reflection}, with much less attention being paid to the other fundamental symmetries: \textit{rotation}, \textit{translation} and \textit{glide reflection}. In the two spatial dimensions relevant for images, these four symmetries can be combined in 17 distinct ways, \textit{the wallpaper groups} \citep{RN1562,RN1563,RN1425}. Previous work on a subset of four of the wallpaper groups used functional MRI to demonstrate that rotation symmetries in wallpapers elicit parametric responses in several areas in occipital cortex, beginning with visual area V3 \citep{RN1725}. This effect was also robust with electroencephalography (EEG), whether measured using Steady-State Visual Evoked Potentials (SSVEPs)\citep{RN1725} or event-related paradigms \citep{RN1959}. Here we extend this work by collecting SSVEPs and psychophysical data from human participants viewing the full set of wallpaper groups. We measure responses in visual cortex to 16 out of the 17 wallpaper groups, with the 17th serving as a control stimulus. Our goal is to provide a more complete picture of how wallpaper groups are represented in the human visual system.

@@

\begin{figure}[hptb]
\centering
\includegraphics[width=0.75\linewidth]{../figures/figure1.pdf}
\caption{\textcolor{ForestGreen}{Subgroup relationships within indices 2 (solid lines) and 3 (dashed line) are shown in (A). All other relationships can be infered by identifying the shortest path through the hierarchy, and multiplying the subgroup indices. For example, \textit{P1} is related to \textit{P6} through \textit{P6}\textrightarrow\textit{P3} (index 2) and \textit{P3}\textrightarrow\textit{P1} (index 3) so \textit{P1} is also a subgroup of \textit{P6} with index $3 \times 2 = 6$. We also show enlarged versions of some of the subgroup relationships involving \textit{PMM} (B, shown in blue) and \textit{P6} (C, shown in red) and highlight the symmetries within the subgroups to emphasize how the supergroup can be generated by adding additional transformations to the subgroup.}}
\end{figure}

@@

The current data provide a description of the visual system’s response to the complete set of symmetries in the two-dimensional plane. Our design precludes us from independently measure the response to \textit{P1}, but because each of the 16 other groups produce non-zero odd harmonic amplitudes (see Figure 2), we can conclude that the relationships between P1 and all other groups, where \textit{P1} is the subgroup, are also preserved by the visual system. The subgroup relationships are in many cases not obvious perceptually, and most participants had no knowledge of group theory. Thus, the visual system’s ability to preserve the subgroup hierarchy does not depend on explicit knowledge of the relationships. \textcolor{ForestGreen}{Previous brain-imaging studies have found evidence of parametric responses with the number of reflection symmetry folds \cite{keefe_2018, Sasaki_2005, makin_2016} and with the order of rotation symmetry \cite{RN1725}. Our study is the first demonstration that the brain encodes symmetry in this parametric fashion across every possible combination of different \textit{symmetry types}, and that this parametric encoding is also reflected in behavior.} Previous behavioral experiments have shown that although naïve observers can distinguish many of the wallpaper groups \citep{RN1253}, they tend to sort them into fewer groups than there actually are (4-12 groups) and it is common for exemplars from different wallpaper groups to be sorted in the same group \citep{RN172}. The more controlled two-interval forced choice approach used in the current behavioral experiment allows us to show that more granular representations of wallpaper groups are measurable in behavior.

\textcolor{ForestGreen}{A large literature exists on the \textit{Sustained Posterior Negativity} (SPN), a characteristic negative-going waveform that is known to reflect responses to symmetry and other forms of regularity and structure \citep{makin_2016}. The SPN scales with the proportion of reflection symmetry in displays that contain a mixture of symmetry and noise \cite{makin_formation_2020, palumbo_scaling_2015}, and both reflection, rotation and translation can produce a measurable SPN \cite{makin_electrophysiological_2013}. It has recently been demonstrated that a holographic model of regularity \citep{van_der_helm_1996}, can predict both SPN amplitude \citep{makin_2016} and perceptual discrimination performance \citep{nucci_2007} for dot patterns that contain symmetry and other types of regularity. The available evidence suggests that the SPN and our SSVEP measurements are two distinct methods of isolating the same symmetry-related brain response: When observed in the time-domain, the symmetry-selective odd-harmonic responses produce similarly sustained waveforms (see Figure 2), odd-harmonic SSVEP responses can be measured for dot patterns similar to those used to measure the SPN \citep{RN1954}, and the one event-related study that has been published on the wallpaper groups also produced SPN-like waveforms \citep{RN1959}. Future work should more firmly establish the connection and determine if the SPN can capture responses similarly precise symmetry responses as the SSVEPs presented here. It would also be worthwhile to ask if and how \textit{W} can computed for our random-noise based wallpaper textures where combinations of symmetries tile the plane.}

We observe a reliable correlation between our brain imaging and psychophysical data. This suggests that the two measurements reflect the same underlying symmetry representations in visual cortex. \textcolor{ForestGreen}{It should be noted that the correlation is relatively modest ($R^2 = 0.44$). This may be partly due to the fact that different individuals participated in the two experiments. It may also be related to the fact that participants where not doing a symmetry-related task during the SSVEP experiment, but instead monitored the stimuli for brief, intermittent contrast changes. Previous brain imaging studies have found enhanced reflection symmetry responses when participants performed a symmetry-related task \cite{makin_formation_2020, Sasaki_2005, keefe_2018}. We did not manipulate task during our SSVEP recordings, but it is possible that adding a symmetry-related task to our SSVEP experiment would have produced measurements that reflected subgroup relationships to an even higher extent. We note, however, that our performance is already close to ceiling (see Figure 5). It is possible that adding a symmetry-related task would enhance SSVEP amplitudes across all wallpaper groups without improving the discriminality of individual groups (similar to what was observed by Keefe and his co-authors \cite{keefe_2018}). SPN measurements suggest that task-driven processing may be important for detecting symmetries that have been subject to perspective distortion \cite{makin_conditions_2015} although it should be noted that this effect was much less clear in a subsequent functional MRI study \cite{keefe_2018}. Future work in which behavioral and brain imaging data are collected from the same participants, and behavior is manipulated in the SSVEP task, will help further establish the connection between the two measurements, and elucidate the potential contribution of task-related top-down processing to the current results.}

@@

\subsection*{Participants}
Twenty-five participants (11 females, mean age 28.7 \textpm\ 3.3) took part in the EEG experiment. Their informed consent was obtained before the experiment under a protocol that was approved by the Institutional Review Board of Stanford University. 11 participants (8 females, mean age 20.73 \textpm\ 1.21) took part in the psychophysics experiment. All participants had normal or corrected-to-normal vision. Their informed consent was obtained before the experiment under a protocol that was approved by the University of Essex's Ethics Committee.

@@

\subsection*{EEG Procedure}
Visual Evoked Potentials were measured using a steady-state design, in which \textit{P1} control images alternated with exemplar images from each of the 16 other wallpaper groups. Exemplar images were always preceded by their matched \textit{P1} control image. A single 0.83 Hz stimulus cycle consisted of a control \textit{P1} image followed by an exemplar image, each shown for 600 ms. A trial consisted of 10 such cycles (12 sec) over which 10 different exemplar images and matched controls from the same rotation group were presented. For each group type, the individual exemplar images were always shown in the same order within the trials. Participants initiated each trial with a button-press, which allowed them to take breaks between trials. Trials from a single wallpaper group were presented in blocks of four repetitions, which were themselves repeated twice per session, and shown in random order within each session. To control fixation, the participants were instructed to fixate a small white cross in the center of display. To control vigilance, a contrast dimming task was employed. Two times per trial, an image pair was shown at reduced contrast, and the participants were instructed to press a button on a response pad. We adjusted the contrast reduction such that average accuracy for each participant was kept at ~85\% correct, in order to keep the difficulty of the vigilance at a constant level.

>>>>

Symmetries are abundant in natural and man-made environments, due to a complex interplay of physical forces that govern pattern formation in nature. Symmetrical patterns have been created and appreciated by human cultures throughout history and since the gestalt movement of the early 20th century, symmetry has been recognized as important for visual perception. Symmetry contributes to the perception of shapes \citep{RN1311,RN1682}, scenes \citep{RN1824} and surface properties \citep{RN1166}, as well as the social process of mate selection \citep{RN1337}. Most of this work has focused on mirror symmetry or \textit{reflection}, with less attention being paid to the other fundamental symmetries: \textit{rotation}, \textit{translation} and \textit{glide reflection} \citep{wagemans_1998}. \textcolor{ForestGreen}{This is perhaps because that reflection has been found to be more perceptually salient \citep{mach_1959,royer_1981,palmer_1991,ogden_2016,hamada_1988} and produced more brain activity \citep{makin_2013,makin_2014,makin_2012,wright_2015} than other types of symmetry. The current study considers all of the four fundamental symmetries not in isolation, but when they are combined in regular textures.} In the two spatial dimensions relevant for images, symmetries can be combined in 17 distinct ways, \textit{the wallpaper groups} \citep{RN1562,RN1563,RN1425}. Previous work on a subset of four of the wallpaper groups used functional MRI to demonstrate that rotation symmetries in wallpapers elicit parametric responses in several areas in occipital cortex, beginning with visual area V3 \citep{RN1725}. This effect was also robust \textcolor{ForestGreen}{when symmetry responses were measured with electroencephalography (EEG) using both Steady-State Visual Evoked Potentials (SSVEPs)\citep{RN1725} and Event-Related Potentials \citep{RN1959}. The SSVEP technique uses periodic visual stimulation to produce a periodic brain response that is confined to integer multiples of the stimulation frequency known as harmonics. SSVEP response harmonics can be isolated in the frequency domain and depending on the specific design, different harmonics will in some cases express different aspects of the brain response. \citep{RN1540}.} Here we extend on the previous work by collecting SSVEPs and psychophysical data from human participants viewing the full set of wallpaper groups. We measure responses in visual cortex to 16 out of the 17 wallpaper groups, with the 17th serving as a control stimulus. Our goal is to provide a more complete picture of how wallpaper groups are represented in the human visual system.

@@

\begin{figure}[hptb]
\centering
\includegraphics[width=0.75\linewidth]{../figures/figure1.pdf}
\caption{\textcolor{ForestGreen}{Subgroup relationships with indices 2 (solid lines) and 3 (dashed line) are shown in (A). All other relationships can be infered by identifying the shortest path through the hierarchy, and multiplying the subgroup indices. For example, \textit{P1} is related to \textit{P6} through \textit{P6}\textrightarrow\textit{P3} (index 2) and \textit{P3}\textrightarrow\textit{P1} (index 3) so \textit{P1} is also a subgroup of \textit{P6} with index $3 \times 2 = 6$. We also show enlarged versions of some of the subgroup relationships involving \textit{PMM} (B, shown in blue) and \textit{P6} (C, shown in red) and highlight the symmetries within the subgroups to emphasize how the supergroup can be generated by adding additional transformations to the subgroup.}}
\end{figure}

@@

\textcolor{ForestGreen}{The current study uses 16 distinct wallpaper groups, while previous neuroimaging studies focused on a subset of 4 \citep{RN1725,RN1959}. This represents a significant conceptual advance, because it makes it possible to investigate the complete subgroup hierarchy among the 17 groups and ask to what extent the hierarchy is reflected in brain activity. Our data} provide a description of the visual system’s response to the complete set of symmetries in the two-dimensional plane. 
\textcolor{ForestGreen}{We do not} independently measure the response to \textit{P1}, but because each of the 16 other groups produce non-zero odd harmonic amplitudes (see Figure 2), we can conclude that the relationships between P1 and all other groups, where \textit{P1} is the subgroup, are also preserved by the visual system. The subgroup relationships are in many cases not obvious perceptually, and most participants had no knowledge of group theory. Thus, the visual system’s ability to preserve the subgroup hierarchy does not depend on explicit knowledge of the relationships. \textcolor{ForestGreen}{Previous brain-imaging studies have found evidence of parametric responses with the number of reflection symmetry folds \cite{keefe_2018, Sasaki_2005, makin_2016} and with the order of rotation symmetry \cite{RN1725}. Our study is the first demonstration that the brain encodes symmetry in this parametric fashion across every possible combination of different \textit{symmetry types}, and that this parametric encoding is also reflected in behavior.} Previous behavioral experiments have shown that although naïve observers can distinguish many of the wallpaper groups \citep{RN1253}, they tend to sort exemplars into fewer \textcolor{ForestGreen}{(4-12) sets than the number of wallpaper groups, often placing exemplars from different wallpaper groups in the same set \citep{RN172}. The two-interval forced choice approach we use in the current psychophysical experiment makes it possible to directly compare symmetry detection thresholds to the subgroup hierarchy, and reveals that not only can the 17 wallpaper groups be distinguished based on behavioral data, behavior largely follows the subgroup hierarchy.}

\textcolor{ForestGreen}{A large literature exists on the \textit{Sustained Posterior Negativity} (SPN), a characteristic negative-going waveform that is known to reflect responses to symmetry and other forms of regularity and structure \citep{makin_2016}. The SPN scales with the proportion of reflection symmetry in displays that contain a mixture of symmetry and noise \cite{makin_formation_2020, palumbo_scaling_2015}, and both reflection, rotation and translation can produce a measurable SPN \cite{makin_2013}. It has recently been demonstrated that a holographic model of regularity \citep{van_der_helm_1996}, can predict both SPN amplitude \citep{makin_2016} and perceptual discrimination performance \citep{nucci_2007} for dot patterns that contain symmetry and other types of regularity. The available evidence suggests that the SPN and our SSVEP measurements are two distinct methods of isolating the same symmetry-related brain response: When observed in the time-domain, the symmetry-selective odd-harmonic responses produce similarly sustained waveforms (see Figure 2), odd-harmonic SSVEP responses can be measured for dot patterns similar to those used to measure the SPN \citep{RN1954}, and the one event-related study that has been published on the wallpaper groups also produced SPN-like waveforms \citep{RN1959}. Future work should more firmly establish the connection and determine if the SPN can capture responses similarly precise symmetry responses as the SSVEPs presented here. It would also be worthwhile to ask if and how \textit{W} can computed for our random-noise based wallpaper textures where combinations of symmetries tile the plane.}

We observe a reliable correlation between our brain imaging and psychophysical data. This suggests that the two measurements reflect the same underlying symmetry representations in visual cortex. \textcolor{ForestGreen}{It should be noted that the correlation is relatively modest ($R^2 = 0.44$). This may be partly due to the fact that different individuals participated in the two experiments. It may also be related to the fact that participants where not doing a symmetry-related task during the SSVEP experiment, but instead monitored the stimuli for brief changes in contrast that occured twice per trial (see Methods). Previous brain imaging studies have found enhanced reflection symmetry responses when participants performed a symmetry-related task \cite{makin_formation_2020, Sasaki_2005, keefe_2018}. We did not manipulate task during our SSVEP recordings, but it is possible that adding a symmetry-related task to our SSVEP experiment would have produced measurements that reflected subgroup relationships to an even higher extent. We note, however, that our performance is already close to ceiling (see Figure 5). It is possible that adding a symmetry-related task would enhance SSVEP amplitudes across all wallpaper groups without improving the discriminality of individual groups (similar to what was observed by Keefe and his co-authors \cite{keefe_2018}). SPN measurements suggest that task-driven processing may be important for detecting symmetries that have been subject to perspective distortion \cite{makin_conditions_2015} although it should be noted that this effect was much less clear in a subsequent functional MRI study \cite{keefe_2018}. Future work in which behavioral and brain imaging data are collected from the same participants, and behavior is manipulated in the SSVEP task, will help further establish the connection between the two measurements, and elucidate the potential contribution of task-related top-down processing to the current results.}

@@

\subsection*{Participants}
Twenty-five participants (11 females, mean age 28.7 \textpm\ 3.3) took part in the EEG experiment. Their informed consent was obtained before the experiment under a protocol that was approved by the Institutional Review Board of Stanford University. 11 participants (8 females, mean age 20.73 \textpm\ 1.21) took part in the psychophysics experiment. All participants had normal or corrected-to-normal vision. Their informed consent was obtained before the experiment under a protocol that was approved by the University of Essex's Ethics Committee. \textcolor{ForestGreen}{There was no overlap in participants between the EEG and psychophysics experiments.}

@@

\subsection*{EEG Procedure}
Visual Evoked Potentials were measured using a steady-state design, in which \textit{P1} control images alternated with exemplar images from each of the 16 other wallpaper groups. Exemplar images were always preceded by their matched \textit{P1} control image. A single 0.83 Hz stimulus cycle consisted of a control \textit{P1} image followed by an exemplar image, each shown for 600 ms. A trial consisted of 10 such cycles (12 sec) over which 10 different exemplar images and matched controls from the same rotation group were presented. For each group type, the individual exemplar images were always shown in the same order within the trials. Participants initiated each trial with a button-press, which allowed them to take breaks between trials. Trials from a single wallpaper group were presented in blocks of four repetitions, which were themselves repeated twice per session, and shown in random order within each session. To control fixation, the participants were instructed to fixate a small white cross in the center of display. To control vigilance, a contrast dimming task was employed. Two times per trial, an image pair \textcolor{ForestGreen}{(control {P1} plus exemplar)} was shown at reduced contrast. \textcolor{ForestGreen}{P}articipants were instructed to press a button on a response pad \textcolor{ForestGreen}{whenever they noticed a contrast change. Reaction times were not taken into account and participants were told to respond at their own pace while being as accurate as possible}. We adjusted the \textcolor{ForestGreen}{reduction in contrast} such that average accuracy for each participant was kept at ~85\% correct, in order to keep the difficulty of the vigilance \textcolor{ForestGreen}{task} at a constant level.
